---
title: "p8105_hw2_lz2949"
author: "Longyi Zhao"
date: "2023-09-27"
output: github_document
---

```{r}
library(tidyverse)
```

## Problem 1
import data
```{r}
pols_df = 
  read_csv("./data/pols-month.csv") |>
  janitor::clean_names() |>
  separate(mon, into = c("year", "month","day"), sep=c(4,7)) |>
  mutate (
    month = 
    case_match(
      month, 
      "-01" ~ "January",
      "-02" ~ "Febuary",
      "-03" ~ "March",
      "-04" ~ "April", 
      "-05" ~ "May",
      "-06" ~ "June",
      "-07" ~ "July",
      "-08" ~ "August",
      "-09" ~ "September",
      "-10" ~ "October",
      "-11" ~ "November",
      "-12" ~ "December",
    )
  ) |>
  mutate (
    president = 
    case_match(
      prez_gop,
    0 ~ "dem", 
    1 ~ "gop", 
    2 ~ "gop"
    )
  ) |>
  subset(select = -c(day, prez_dem, prez_gop))
```

clean data in snp.csv
```{r}
snp_df = 
  read_csv("./data/snp.csv") |>
  janitor::clean_names() |> 
  separate(date, into = c("month", "day", "year"), sep = "/") |> # drop day
  subset(select = -day) |>
  mutate (
    month = 
    case_match(
      month, 
      "1" ~ "January",
      "2" ~ "Febuary",
      "3" ~ "March",
      "4" ~ "April", 
      "5" ~ "May",
      "6" ~ "June",
      "7" ~ "July",
      "8" ~ "August",
      "9" ~ "September",
      "10" ~ "October",
      "11" ~ "November",
      "12" ~ "December",
    ))
# add 19 when num(year) is greater than 19, else add 20
snp_df$year = ifelse(as.numeric(snp_df$year)>= 50, paste0("19",snp_df$year), paste0("20", snp_df$year))
# switch the order between year and month 
snp_df = snp_df[, c("year", "month", "close")]
```

tidy up unemployment data 
```{r}
unemp_df = 
  read_csv("./data/unemployment.csv") |>
  janitor::clean_names() |> # switching from wide to long 
  pivot_longer(jan:dec,
               names_to = "month",
               values_to = "unemployment rate") |>
  mutate (
    year = as.character(year),
    month = 
    case_match(
      month, 
      "jan" ~ "January",
      "feb" ~ "Febuary",
      "mar" ~ "March",
      "apr" ~ "April", 
      "may" ~ "May",
      "jun" ~ "June",
      "jul" ~ "July",
      "aug" ~ "August",
      "sep" ~ "September",
      "oct" ~ "October",
      "nov" ~ "November",
      "dec" ~ "December",
    ))
```

join the datasets by merging snp into pols, and merging unemployment into results

left-join snp into pols, then add unemployment data as well

```{r}
snp_pols_data = 
  left_join(pols_df, snp_df, by = c("year", "month"))

merged_data = left_join(snp_pols_data, unemp_df, by = c("year", "month"))
```


write a short paragraph about these datasets. explain briefly what each dataset contained, and describe the resulting dataset (dimention, range of years, names of key variables)

## Problem 2

```{r}
library(readxl) # read Mr. Trash Wheel data
MtrashW_df = 
  read_excel("./data/202207 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N547") |>
  janitor::clean_names() |>
  mutate(
    homes_powered = (
      weight_tons*500/30
    )
  )

MtrashW_df$Name = "Mr. Trash Wheel"


```

use similar process to import, clean, and organize data for Professor Trash Wheel and Gwynnda. 

```{r}
PtrashW_df = 
  read_excel("./data/202207 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = "A2:M96") |>
  janitor::clean_names() |>
    mutate(
    year = as.character(year),
    homes_powered = (
      weight_tons*500/30
    )
  )
PtrashW_df$Name = "Professor Trash Wheel"


Gwynnda_df = 
  read_excel("./data/202207 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda Trash Wheel", range = "A2:K108") |>
  janitor::clean_names() |>
    mutate(
    year = as.character(year),
    homes_powered = (
      weight_tons*500/30
    )
  )
Gwynnda_df$Name = "Gwynnda Trash Wheel"
```

Combine these two dataset with Mr. Trash Wheel. Use the bind_rows ?

```{r}
trash_tidy = 
  bind_rows (MtrashW_df, PtrashW_df, Gwynnda_df) |>
  janitor::clean_names() 
# move trash wheel name to the first column 
trash_tidy = trash_tidy[, c("name", setdiff(names(trash_tidy), "name"))]
```

write a paragraph about these data

## Problem 3

import, clean, and tidy baseline demographic database

```{r}
baseline_df = 
  read_csv("./data_mci/MCI_baseline.csv", skip = 1) |>
  janitor::clean_names() |> # convert sex/apoe4 from number to name 
  mutate (
    sex = case_match(
      sex,
      1 ~ "male", 
      0 ~ "female"
    ),
    apoe4 = case_match(
      apoe4,
      1 ~ "APOE4 carrier",
      0 ~ "APOE4 non-carrier"
    )
  ) 
  

```


discuss the importatn steps in the import process and relevant features of the dataset

Import, clean and tidt longitudinally dataset

```{r}
amyloid_df = 
  read_csv("./data_mci/mci_amyloid.csv", skip = 1) |>
  janitor::clean_names() 
```

comment on steps on the import process and features of the dataset

check whether some participants appear in only the baseline or amyloid dataset











