---
title: "p8105_hw2_lz2949"
author: "Longyi Zhao"
date: "2023-09-27"
output: github_document
---

```{r}
library(tidyverse)
```

## Problem 1
Import data
```{r}
pols_df = 
  read_csv("./data/pols-month.csv") |>
  janitor::clean_names() |>
  separate(mon, into = c("year", "month","day"), sep=c(4,7)) |>
  mutate (
    month = 
    case_match(
      month, 
      "-01" ~ "January",
      "-02" ~ "Febuary",
      "-03" ~ "March",
      "-04" ~ "April", 
      "-05" ~ "May",
      "-06" ~ "June",
      "-07" ~ "July",
      "-08" ~ "August",
      "-09" ~ "September",
      "-10" ~ "October",
      "-11" ~ "November",
      "-12" ~ "December",
    )
  ) |>
  mutate (
    president = 
    case_match(
      prez_gop,
    0 ~ "dem", 
    1 ~ "gop", 
    2 ~ "gop"
    )
  ) |>
  subset(select = -c(day, prez_dem, prez_gop))
```

Clean data in snp.csv
```{r}
snp_df = 
  read_csv("./data/snp.csv") |>
  janitor::clean_names() |> 
  separate(date, into = c("month", "day", "year"), sep = "/") |> # drop day
  subset(select = -day) |>
  mutate (
    month = 
    case_match(
      month, 
      "1" ~ "January",
      "2" ~ "Febuary",
      "3" ~ "March",
      "4" ~ "April", 
      "5" ~ "May",
      "6" ~ "June",
      "7" ~ "July",
      "8" ~ "August",
      "9" ~ "September",
      "10" ~ "October",
      "11" ~ "November",
      "12" ~ "December",
    ))
# add 19 when num(year) is greater than 19, else add 20
snp_df$year = ifelse(as.numeric(snp_df$year)>= 50, paste0("19",snp_df$year), paste0("20", snp_df$year))
# switch the order between year and month 
snp_df = snp_df[, c("year", "month", "close")]
```

Tidy up unemployment data 
```{r}
unemp_df = 
  read_csv("./data/unemployment.csv") |>
  janitor::clean_names() |> # switching from wide to long 
  pivot_longer(jan:dec,
               names_to = "month",
               values_to = "unemployment rate") |>
  mutate (
    year = as.character(year),
    month = 
    case_match(
      month, 
      "jan" ~ "January",
      "feb" ~ "Febuary",
      "mar" ~ "March",
      "apr" ~ "April", 
      "may" ~ "May",
      "jun" ~ "June",
      "jul" ~ "July",
      "aug" ~ "August",
      "sep" ~ "September",
      "oct" ~ "October",
      "nov" ~ "November",
      "dec" ~ "December",
    ))
```

Join the datasets by merging snp into pols, and merging unemployment into results

Left-join snp into pols, then add unemployment data as well

```{r}
snp_pols_data = 
  left_join(pols_df, snp_df, by = c("year", "month"))

merged_data = left_join(snp_pols_data, unemp_df, by = c("year", "month"))
```

The merged data has `r nrow(merged_data)` rows and `r ncol(merged_data)` columns. It contains three datasets. The first one is called pols_df, it contains `r nrow(pols_df)` rows and `r ncol(pols_df)` columns data, ranging from `r range(pols_df$year)[1]` to `r range(pols_df$year)[2]`. The data is about number of republican/democratic governors/senators on associated date and whether the president is from republican or democratic. The second one is called snp_df, it contains `r nrow(snp_df)` rows and `r ncol(snp_df)` columns data, ranging from `r range(snp_df$year)[1]` to `r range(snp_df$year)[2]`. It described closing values of the stock index on the associated date. The third dataset is called unemp_df, it contains `r nrow(unemp_df)` rows and `r ncol(unemp_df)` columns data, ranging from `r range(unemp_df$year)[1]` to `r range(unemp_df$year)[2]`. It described the percentage of unemployment in given month-year. 

## Problem 2

```{r}
library(readxl) # read Mr. Trash Wheel data
MtrashW_df = 
  read_excel("./data/202309 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N586") |>
  janitor::clean_names() |>
  mutate(
    homes_powered = (
      weight_tons*500/30
    )
  )

MtrashW_df$Name = "Mr. Trash Wheel"


```


```{r}
PtrashW_df = 
  read_excel("./data/202309 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = "A2:M108") |>
  janitor::clean_names() |>
    mutate(
    year = as.character(year),
    homes_powered = (
      weight_tons*500/30
    )
  )
PtrashW_df$Name = "Professor Trash Wheel"


Gwynnda_df = 
  read_excel("./data/202309 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda Trash Wheel", range = "A2:L157") |>
  janitor::clean_names() |>
    mutate(
    year = as.character(year),
    homes_powered = (
      weight_tons*500/30
    )
  )
Gwynnda_df$Name = "Gwynnda Trash Wheel"
```

Combine these two dataset with Mr. Trash Wheel. 

```{r}
trash_tidy = 
  bind_rows (MtrashW_df, PtrashW_df, Gwynnda_df) |>
  janitor::clean_names() 
# move trash wheel name to the first column 
trash_tidy = trash_tidy[, c("name", setdiff(names(trash_tidy), "name"))]
```

There are three datasets in this problem. The first one is called Mr. Trash Wheel, it has `r length(names(MtrashW_df))` variables and it includes data of `r nrow(MtrashW_df)` dumpsters, ranging from `r range(MtrashW_df$year)[1]` to `r range(MtrashW_df$year)[2]`. The second one is called Professor Trash Wheel, it has `r length(names(PtrashW_df))` variables and it includes data of `r nrow(PtrashW_df)` dumpsters, ranging from `r range(PtrashW_df$year)[1]` to `r range(PtrashW_df$year)[2]`. Similarly, the third one is called Gwynnda Trash Wheel, it has `r length(names(Gwynnda_df))` variables and it includes data of `r nrow(Gwynnda_df)` dumpsters, ranging from `r range(Gwynnda_df$year)[1]` to `r range(Gwynnda_df$year)[2]`. Some of their collected trash are `r names(trash_tidy)[8:10]`. 

The total weight of trash collected by Professor Trash Wheel is `r sum(PtrashW_df$weight_tons)` tons. The total number of cigarette butts collected by Gwynnda in July of 2021 is `r sum(subset(Gwynnda_df, month == "July" & year == "2021")$cigarette_butts)`. 


## Problem 3

Import, clean, and tidy baseline demographic database

```{r}
baseline_df = 
  read_csv("./data_mci/MCI_baseline.csv", skip = 1) |>
  janitor::clean_names() |> # convert sex/apoe4 from number to name 
mutate ( 
    sex = case_match(
      sex,
      1 ~ "male", 
      0 ~ "female"
    ),
    apoe4 = case_match(
      apoe4,
      1 ~ "APOE4 carrier",
      0 ~ "APOE4 non-carrier"
    )
  ) # people who develop MCI should not have MCI at the begining 
c_baseline_df = filter(baseline_df, age_at_onset > current_age | age_at_onset == ".")
```

Description: To import the baseline dataset, read_csv function is used and skip = 1 is used to skip the explanation line above header. Mutate function is then used to change sex values to male or female and apoe4 values to carrier or non-carrier. Finally, the filter function is used to filter out those who do not have MCI values recorded. 

The total number of participants is `r nrow(baseline_df)`. Among those, `r nrow(c_baseline_df)` does not have MCI at the begining of the study. Later, `r nrow(filter(c_baseline_df, age_at_onset != "."))` developed MCI. The average baseline age of those `r nrow(filter(c_baseline_df, age_at_onset != "."))` patients is `r round(mean(as.numeric(filter(c_baseline_df, age_at_onset != ".")$current_age)), digits = 2)`. 

In this study, `r nrow(subset(c_baseline_df, sex == "female" & apoe4 == "APOE4 carrier"))` female patients are apoe4 carrier, which counts for `r round(nrow(subset(c_baseline_df, sex == "female" & apoe4 == "APOE4 carrier"))/nrow(subset(c_baseline_df, sex == "female"))*100, digits = 2)`% proportion of total female participants meeting the inclusion criteria. 


Import, clean and tidy longitudinally dataset. 
```{r}
amyloid_df = 
  read_csv("./data_mci/mci_amyloid.csv", skip = 1) |>
  janitor::clean_names() |>
  rename(id = study_id) # change study_id to id to match column names to baseline_df
```

Read_csv is also used to import dataset of longitudinally observed biomarker values. To help the comparison between two datasets easier, the column name study_id is changed to id using rename(). In this longitudinal study, there are total `r nrow(amyloid_df)` participants being traced. 

```{r}
amyloid_df = 
  pivot_longer(
    amyloid_df, 
    baseline:time_8, 
    names_to = "time",
    values_to = "time_elapse"
  )
```

Pivot longer is then used to make the data looks more tidy. 
The Amyloid dataset contains time (in years) elapsed since the study baseline to the visit where biomarker Amyloid: 42/40 ratio was measured. There are total `r as.numeric(nrow(amyloid_df))/5` patients being measured. 

```{r}
same_id_baseline = sum(baseline_df$id %in% amyloid_df$id) # check the number of id in baseline apper in amyloid
same_id_c_baseline = sum(c_baseline_df$id %in% amyloid_df$id)
```

In amyloid and baseline datasets, `r same_id_baseline` appeared in both, some of them are healthy while `r same_id_c_baseline` of them developed MCI later. 

Combine two dataset so who appear in both datasets are retained 
```{r}
m_amybase = merge(c_baseline_df, amyloid_df, by = "id")
```

Values from two datasets of those `r same_id_c_baseline` patients are combined. `r nrow(subset(m_amybase, sex == "female"))` of them are female and `r nrow(subset(m_amybase, sex == "male"))` of them are male. `r nrow(subset(m_amybase, sex == "male" & apoe4 == "APOE4 carrier"))` males are APOE4 carriers and `r nrow(subset(m_amybase, sex == "female" & apoe4 == "APOE4 carrier"))` females are APOE4 carriers. 

Export combined dataset as CSV to the data_mci directory
```{r}
write_csv(m_amybase, "data_mci/m_amybase.csv")
```







